
\subsection{Ablation Studies}
\label{sec:ablations}

We ablate the two quality gates in the EmToM generation pipeline to measure each gate's independent contribution.
Both ablations generate 50 tasks targeting GPT-5.4 at hard difficulty with ICL-guided generation.
After generation, we retroactively apply the removed gate to measure what it would have caught.

\subsubsection{Ablation 1: Removing the Baseline Solvability Check}
\label{sec:ablation_baseline}

The benchmark gate normally runs the target model in two modes: \emph{standard} (partial information) and \emph{baseline} (full information), accepting only tasks where standard fails but baseline succeeds.
This ensures difficulty comes from information asymmetry---not from the task being fundamentally unsolvable.

In this ablation, we remove the baseline run: the gate accepts any task where standard mode fails, regardless of whether baseline would succeed.
The judge gate still runs normally.

We then run the baseline benchmark post-hoc on all generated tasks.

\begin{table}[h]
\centering
\caption{Ablation 1: Post-hoc baseline results on 49 tasks generated without the baseline solvability check.}
\label{tab:ablation_baseline}
\begin{tabular}{@{}lcc@{}}
\toprule
\textbf{Post-hoc baseline outcome} & \textbf{Count} & \textbf{\%} \\
\midrule
Baseline fails (no functional ToM signal) & 25/49 & 51\% \\
Baseline passes (functional ToM signal) & 24/49 & 49\% \\
\bottomrule
\end{tabular}
\end{table}

Without the baseline check, 51\% of tasks are unsolvable even with full information---they test general task complexity rather than Theory of Mind.

\subsubsection{Ablation 2: Removing the LLM Council Judge}
\label{sec:ablation_judge}

The judge gate scores each task on 8 quality criteria using a two-model council.
A task passes only if its overall score is $\geq 0.65$ and every individual criterion scores $\geq 0.5$.

In this ablation, we remove the judge entirely: the benchmark gate (standard + baseline) is the only quality filter.
We then run the LLM council post-hoc on all generated tasks.

\begin{table}[h]
\centering
\caption{Ablation 2: Post-hoc judge results on 50 tasks generated without the LLM council.}
\label{tab:ablation_judge}
\begin{tabular}{@{}lccc@{}}
\toprule
\textbf{Council models} & \textbf{Fail rate} & \textbf{Avg.\ score} & \textbf{Min score} \\
\midrule
kimi-k2.5 + GPT-5.2 & 50\% & 0.81 & 0.54 \\
kimi-k2.5 + GPT-5.4 & 90\% & 0.75 & 0.39 \\
\bottomrule
\end{tabular}
\end{table}

Without the judge, 50--90\% of tasks fail quality checks depending on the council model strength.
A stronger judge model (GPT-5.4) catches more issues, primarily: mechanics that are present but not load-bearing, secrets that leak exact object IDs, and mixed tasks whose personal goals duplicate rather than conflict with shared goals.
